\pkg{plotROC}: A Tool for Plotting ROC Curves

### 2016-07-15 Paper/sachs2015plotROC.Rmd
revision number two

---
title: '\pkg{plotROC}: A Better Tool for Plotting ROC Curves'
shorttitle: '\pkg{plotROC}: A Better Tool for Plotting ROC Curves'
plaintitle: 'plotROC: A Better Tool for Plotting ROC Curves'
author:
- name: Michael C. Sachs
  affil: Biometric Research Branch, Division of Cancer Treatment and Diagnosis, National Cancer Institute
address:
- Michael C. Sachs
- 9609 Medical Center Drive, MSC 9735
- Bethesda, MD 20892
- "telephone: 240-276-7888"
email: michael.sachs@nih.gov
abstract: 'Plots of the receiver operating characteristic (ROC) curve are ubiquitous in medical research. Designed to simultaneously display the operating characteristics at every possible value of a continuous diagnostic test, ROC curves are used in oncology to evaluate screening, diagnostic, prognostic and predictive biomarkers. I reviewed a sample of ROC curve plots from the major oncology journals in order to assess current trends in usage and design elements. My review suggests that ROC curve plots are often ineffective as statistical charts and that poor design obscures the relevant information the chart is intended to display. I describe my new \proglang{R} package that was created to address the shortcomings of existing tools. The package has functions to create informative ROC curve plots, with sensible defaults and a simple interface, for use in print or as an interactive web-based plot. A web application was developed to reach a broader audience of scientists who do not use \proglang{R}.'
keywords: ROC curves; graphics; interactive; plots
bibliography: plotroc
...

@@

Given a sample of test and disease status pairs, $(M_1, D_1), \ldots, (M_n, D_n)$, I can estimate the ROC curve by computing proportions in the diseased and healthy subgroups separately. Specifically, given a fixed cutoff $c$, an estimate of the $TPF(c)$ is the proportion of test positives among the diseased subgroup:

@@

where $1\{\cdot\}$ is the indicator function that equals 1 when the condition inside the parentheses is true and 0 otherwise. An estimate for $FPF(c)$ is the proportion of test positives among the healthy subgroup and is given by a similar expression with $D_i = 1$ replaced with $D_i = 0$. Calculating these proportions for $c$ equal to each unique value of the observed $M_i$ yields what is known as the empirical ROC curve estimate. The empirical estimate is a step function. Other methods exist to estimate the ROC curve, such as the binormal parametric estimate which can be used to get a smooth curve. There are also extensions that allow for estimation with time-to-event outcomes subject to censoring. For a more thorough reference on the methods and theory surrounding ROC curves, I refer interested readers to @pepe2003statistical.

@@

In the medical literature, FPF and TPF are usually referred to in terms of the jargon sensitivity and specificity. Sensitivity is equivalent to the true positive fraction. Specificity is 1 - FPF, the true negative fraction. Sometimes, the FPF and TPF are incorrectly referred to as rates, using the abbreviations FPR and TPR. These are probabilities and their estimates are proportions, therefore I prefer the use of the term fraction as opposed to rate.

@@

The ROC curve plot is, at the most basic level, a line graph. Therefore, once the appropriate statistics are estimated, existing plotting functions can be used to create an ROC curve plot. Viewers can identify ROC plots through context, by observing the shape of the line, and through the addition of axis labels, titles, legends, and so on. In my review of the oncology literature, I observed plots with the distinctive characteristics of the plotting functions from Microsoft Excel [@excel], \proglang{SAS} [@sas], \proglang{SPSS} [@spss], and the base \proglang{R} plotting functions [@arr].

There are several \proglang{R} packages related to ROC curve estimation that contain dedicated plotting functions. The \pkg{ROCR} package [@rocr] plots the FPF _versus_ TPF, as usual, and then takes the interesting approach of encoding the cutoff values as a separate color scale along the ROC curve itself. A legend for the color scale is placed along the vertical axis on the right of the plotting region. The \pkg{pROC} package [@pROC] provides an option for plotting cutoff labels (\code{print.thres = TRUE}) and is mainly focused on estimating confidence intervals and regions for restricted ranges of the ROC curve. The plotting methods therein use the base \proglang{R} plotting functions to create nice displays of the curves along with shaded confidence regions. My \pkg{plotROC} package uses the \pkg{ggplot2} [@ggplot2] plotting library to create clear, informative ROC plots, with interactive features for use on the web, and sensible defaults for use in print.

@@

The panels of Figure \ref{figure1} illustrate the most common styles of ROC curve plots, and the associated design elements. I favor the use of gridlines and a diagonal reference line to facilitate accurate readings off of the axes. Direct labels are preferred over legends because they omit the additional cognitive step of matching line types or colors to labels. My \pkg{plotROC} package additionally provides plotting of cutoff values, which are displayed interactively with the web-based output option, and direct labels for print use. Exact confidence regions for points on the ROC curve are optionally calculated and displayed. Additionally, I use axis scales adjusted to be denser near the margins 0 and 1. In medical applications, it is often necessary to have a very low FPR (less than 10%, for instance), therefore the smaller scales are useful for accurately determining values near the margins. The next section details the usage of the \pkg{plotROC} \proglang{R} package and these features.

@@

The vast majority of the figures that I reviewed looked more like those in panels A and B than those in C and D. While plots like this do technically display the trade-offs between false- and true-positives, they minimize the amount of useful information that can be displayed in the space needed to plot an ROC curve. The plots created by \pkg{plotROC} attempt to increase the amount of information displayed in ROC charts that would otherwise be mostly white space. This is useful not only for print media, where space is limited, but also during data analysis. The analyst can quickly and easily view information that would otherwise be obscured by standard plotting software. The interactive features take this one step further, enhancing the plots with high density and easily accessible supplementary information.

@@

I created a \pkg{shiny} application [@shiny] in order to make the features more accessible to non-\proglang{R} users. A limited subset of the functions of \pkg{plotROC} can be performed on an example dataset or on data that users upload to the website. Resulting plots can be saved to the users' machine as a pdf or as a stand-alone html file.  It can be used in any modern web browser with no other dependencies at the website here: http://sachsmc.shinyapps.io/plotROC.

@@

Users can install the latest version of \pkg{plotROC} from github, as shown below. The package will appear on the Comprehensive \proglang{R} Archive Network in the near future. It can also be downloaded from https://github.com/sachsmc/plotROC and installed from source. As of October 19, 2015, \pkg{plotROC} also requires the development version of \pkg{ggplot2} also available from github.

```{r load, eval = FALSE}
devtools::install_github("hadley/ggplot2")
devtools::install_github("sachsmc/plotROC")
library(plotROC)
```

@@

As of version 1.0.1, the \pkg{ggplot2} package [@ggplot2] exports the previously internal functions \code{Geom} and \code{Stat}. This enables developers to create their own statistical tranformations (stats) and geometric layers (geoms), while enjoying all of the other features of \pkg{ggplot2}. We have implemented the empirical ROC curve estimate and the calculation of exact confidence regions as statistical transformations: \code{stat_roc} and \code{stat_rocci}, respectively. We have also defined geometric layers for the ROC curve and confidence regions for the ROC curve: \code{geom_roc} and \code{geom_rocci}, respectively. For further discussion and details of the grammar of graphics as implemented in \pkg{ggplot2}, we refer readers to @wickham2010layered and the \pkg{ggplot2} vignettes.

To use the ROC geometric layer, I use the \code{ggplot} function to define the aesthetic mappings, and the \code{geom_roc} function to add an ROC curve layer. The \code{geom_roc} function requires that the named aesthetics \code{d} for disease status, and \code{m} for marker be present in the \code{aes} function call inside \code{ggplot}. By default, the ROC geom and stat are linked, so that when \code{geom_roc} is called, \code{stat_roc} does the computation, and when \code{stat_roc} is called, \code{geom_roc} is used to plot the layer. The disease status need not be coded as 0/1, but if it is not, \code{stat_roc} assumes (with a warning) that the lowest value in sort order signifies disease-free status.

@@

The same objects like \code{basicplot} with Roc and/or Rocci layers can be treated like any other ggplot objects. It can be printed to display the figure, and other layers can be added to the plot. We provide the function `style_roc()` which is a layer containing a theme, modified gridlines, and axes. Adding the `style_roc()` layer to the ggplot object creates a plot with sensible defaults for use in print. This function has options for the number and location of major and minor breaks, addition of the diagonal guideline, axis labels, and any theme object created for use with ggplot2 can be supplied.

@@

Hovering over the display shows the cutoff value at the point nearest to the cursor. Clicking makes the cutoff label stick until the next click, and if confidence regions are available, clicks will also display those as grey rectangles. The confidence regions are automatically detected. When the user clicks on the ROC curve, the confidence region for the TPF and FPF is overlaid using a grey rectangle. The label and region stick until the next click.

@@

By default \code{calculate_roc} computes the empirical ROC curve. There are other estimation methods out there, as I have summarized in the introduction.  Any estimation method can be used, as long as the cutoff, the TPF and the FPF are returned. Then you can simply pass those values in a data frame to the \code{ggplot} function and override the default statistical tranformation to \code{stat_identity}. For example, let us use the binormal method to create a smooth curve. This approach assumes that the test distribution is normal conditional on disease status.

@@

\pkg{plotROC} makes interactive plots by first converting the \code{ggplot} object into a scalable vector graphic (svg) object with the \code{gridSVG::grid.export} function. This function maps each element of the plot to a corresponding element of the svg markup language. I keep track of the names of the points and labels elements so that I can add interactivity using \pkg{d3.js} and \proglang{JavaScript}. The main interactive feature I wanted was to be able to display the cutoff labels at the points on the ROC curve closest to the mouse cursor.

There are many ways to solve this with \pkg{d3.js}, but I decided to use Voronoi polygons to map the cursor location to the nearest point on the ROC curve. The idea is that for the set of cutoff points along the ROC curve, the \code{d3.geom.voronoi} function chain computes a set of polygons overlaying the plotting region such that the area of each polygon contains the region of the plot closest to it's corresponding cutoff point. Hover events are bound to the polygons so that when the mouse cursor moves around the plotting region, the closest point on the ROC curve is made visible. Similarly, click events are bound to the polygons so that the appropriate confidence region is made visible upon clicking. The svg code and all necessary \proglang{JavaScript} code is returned in the character string provided by \code{export_interactive_roc}.

This approach is similar to what is done in the \pkg{gridSVG} \code{grid.animate} function, which uses the svg \code{<animate />} tags. However, the available features were not sufficient for my needs, which is why I used \pkg{d3.js}. There are several other \proglang{R} packages that aim to create interactive figures. The authors of \pkg{animint} [@animint] created an extensive \proglang{JavaScript} library that creates plots in a similar way as \pkg{ggplot2}. A set of interactive features can be added to plots using \pkg{d3.js}. \pkg{ggvis} [@ggvis], \pkg{rCharts} [@rcharts], and the more recently released \pkg{htmlwidgets} [@htmlwidgets] all leverage existing charting libraries written in \proglang{JavaScript}. \pkg{qtlcharts} [@qtlcharts] uses a set of custom \proglang{JavaScript} and \pkg{d3.js} functions to visualize data from genetic experiments. Their general approach is to manipulate the data and create options in \proglang{R}, and then let the charting libraries or functions handle the rendering and interactivity. \pkg{plotROC} lets \proglang{R} do the rendering, allowing the figures to be consistent across print and web-based media, and retaining the distinctive \proglang{R} style. This also allows users to manipulate the figures directly in \proglang{R} to suit their needs, using tools that are more accessible and familiar to most \proglang{R} users. Then, the \proglang{JavaScript} adds a layer of interactivity to the rendered figures.

>>>>

---
title: '\pkg{plotROC}: A Tool for Plotting ROC Curves'
shorttitle: '\pkg{plotROC}: A Tool for Plotting ROC Curves'
plaintitle: 'plotROC: A Tool for Plotting ROC Curves'
author:
- name: Michael C. Sachs
  affil: Biometric Research Branch, Division of Cancer Treatment and Diagnosis, National Cancer Institute
address:
- Michael C. Sachs
- 9609 Medical Center Drive, MSC 9735
- Bethesda, MD 20892
- "telephone: 240-276-7888"
email: michael.sachs@nih.gov
abstract: 'Plots of the receiver operating characteristic (ROC) curve are ubiquitous in medical research. Designed to simultaneously display the operating characteristics at every possible value of a continuous diagnostic test, ROC curves are used in oncology to evaluate screening, diagnostic, prognostic and predictive biomarkers. I reviewed a sample of ROC curve plots from the major oncology journals in order to assess current trends in usage and design elements. My review suggests that ROC curve plots are often ineffective as statistical charts and that poor design obscures the relevant information the chart is intended to display. I describe my new \proglang{R} package that was created to address the shortcomings of existing tools. The package has functions to create informative ROC curve plots, with sensible defaults and a simple interface, for use in print or as an interactive web-based plot. A web application was developed to reach a broader audience of scientists who do not use \proglang{R}.'
keywords: ROC curves; graphics; interactive; plots
bibliography: plotroc
...

@@

Given a sample of test and disease status pairs, $(M_1, D_1), \ldots, (M_n, D_n)$, one can estimate the ROC curve by computing proportions in the diseased and healthy subgroups separately. Specifically, given a fixed cutoff $c$, an estimate of the $TPF(c)$ is the proportion of test positives among the diseased subgroup:

@@

where $1\{\cdot\}$ is the indicator function that equals 1 when the condition inside the parentheses is true and 0 otherwise. An estimate for $FPF(c)$ is the proportion of test positives among the healthy subgroup and is given by a similar expression with $D_i = 1$ replaced with $D_i = 0$. Calculating these proportions for $c$ equal to each unique value of the observed $M_i$ yields what is known as the empirical ROC curve estimate. The empirical estimate is a step function. Other methods exist to estimate the ROC curve, such as the binormal parametric estimate which can be used to get a smooth curve. There are also extensions that allow for estimation with time-to-event outcomes subject to censoring. For a more thorough reference on the methods and theory surrounding ROC curves, interested readers may look to @pepe2003statistical.

@@

In the medical literature, FPF and TPF are usually referred to in terms of the jargon sensitivity and specificity. Sensitivity is equivalent to the true positive fraction. Specificity is 1 - FPF, the true negative fraction. Sometimes, the FPF and TPF are incorrectly referred to as rates, using the abbreviations FPR and TPR. These are probabilities and their estimates are proportions, therefore the use of the term fraction as opposed to rate is preferred.

@@

The ROC curve plot is, at the most basic level, a line graph. Therefore, once the appropriate statistics are estimated, existing plotting functions can be used to create an ROC curve plot. Viewers can identify ROC plots through context, by observing the shape of the line, and through the addition of axis labels, titles, legends, and so on. In a review of the oncology literature, I observed plots with the distinctive characteristics of the plotting functions from Microsoft Excel [@excel], \proglang{SAS} [@sas], \proglang{SPSS} [@spss], and the base \proglang{R} plotting functions [@arr].

There are several \proglang{R} packages related to ROC curve estimation that contain dedicated plotting functions. The \pkg{ROCR} package [@rocr] plots the FPF _versus_ TPF, as usual, and then takes the interesting approach of encoding the cutoff values as a separate color scale along the ROC curve itself. A legend for the color scale is placed along the vertical axis on the right of the plotting region. The \pkg{pROC} package [@pROC] provides an option for plotting cutoff labels (\code{print.thres = TRUE}) and is mainly focused on estimating confidence intervals and regions for restricted ranges of the ROC curve. The plotting methods therein use the base \proglang{R} plotting functions to create nice displays of the curves along with shaded confidence regions. The \pkg{plotROC} package uses the \pkg{ggplot2} [@ggplot2] plotting package to create clear, informative ROC plots, with interactive features for use on the web, and sensible defaults for use in print.

@@

The panels of Figure \ref{figure1} illustrate the most common styles of ROC curve plots, and the associated design elements. I favor the use of gridlines and a diagonal reference line to facilitate accurate readings off of the axes. Direct labels are preferred over legends because they omit the additional cognitive step of matching line types or colors to labels. The \pkg{plotROC} package additionally provides plotting of cutoff values, which are displayed interactively with the web-based output option, and direct labels for print use. Exact confidence regions for points on the ROC curve are optionally calculated and displayed. Additionally, the use of axis scales adjusted to be denser near the margins 0 and 1 facilitates accurate reading of FPF and TPF values. In medical applications, it is often necessary to have a very low FPF (less than 10%, for instance), therefore the smaller scales are useful for accurately determining values near the margins. The next section details the usage of the \pkg{plotROC} \proglang{R} package and these features.

@@

The vast majority of the figures that were reviewed looked more like those in panels A and B than those in C and D. While plots like this do technically display the trade-offs between false- and true-positives, they minimize the amount of useful information that can be displayed in the space needed to plot an ROC curve. The plots created by \pkg{plotROC} attempt to increase the amount of information displayed in ROC charts that would otherwise be mostly white space. This is useful not only for print media, where space is limited, but also during data analysis. The analyst can quickly and easily view information that would otherwise be obscured by standard plotting software. The interactive features take this one step further, enhancing the plots with high density and easily accessible supplementary information.

@@

A \pkg{shiny} application [@shiny] was created in order to make the features more accessible to non-\proglang{R} users. A limited subset of the functions of \pkg{plotROC} can be performed on an example dataset or on data that users upload to the website. Resulting plots can be saved to the users' machine as a pdf or as a stand-alone html file.  It can be used in any modern web browser with no other dependencies at the website here: http://sachsmc.shinyapps.io/plotROC.

@@

Users can install the latest version of \pkg{plotROC} from the Comprehensive \proglang{R} Archive Network (CRAN). It can also be downloaded or installed from https://github.com/sachsmc/plotROC, where active development will take place. \pkg{plotROC} also requires at least version 2.0.1 of \pkg{ggplot2} which is also available from CRAN.

```{r load, eval = FALSE}
install.packages(c("ggplot2", "plotROC"))
library(plotROC)
```

@@

As of version 2.0.1, the \pkg{ggplot2} package [@ggplot2] exports the previously internal functions \code{Geom} and \code{Stat}. This enables developers to create their own statistical tranformations (stats) and geometric layers (geoms), while enjoying all of the other features of \pkg{ggplot2}. I have implemented the empirical ROC curve estimate and the calculation of exact confidence regions as statistical transformations: \code{stat_roc} and \code{stat_rocci}, respectively. I have also defined geometric layers for the ROC curve and confidence regions for the ROC curve: \code{geom_roc} and \code{geom_rocci}, respectively. For further discussion and details of the grammar of graphics as implemented in \pkg{ggplot2}, I refer readers to @wickham2010layered and the \pkg{ggplot2} vignettes.

To use the ROC geometric layer, I use the \code{ggplot} function to define the aesthetic mappings, and the \code{geom_roc} function to add an ROC curve layer. The \code{geom_roc} function requires that the named aesthetics \code{d} for disease status, and \code{m} for marker, be present in the \code{aes} function call inside \code{ggplot}. By default, the ROC geom and stat are linked, so that when \code{geom_roc} is called, \code{stat_roc} does the computation, and when \code{stat_roc} is called, \code{geom_roc} is used to plot the layer. The disease status need not be coded as 0/1, but if it is not, \code{stat_roc} assumes (with a warning) that the lowest value in sort order signifies disease-free status.

@@

The same objects like \code{basicplot} with Roc and/or Rocci layers can be treated like any other ggplot objects. They can be printed to display the figure, and other layers can be added to the plot. I provide the function `style_roc()` which is a layer containing a theme, modified gridlines, and axes. Adding the `style_roc()` layer to the ggplot object creates a plot with sensible defaults for use in print. This function has options for the number and location of major and minor breaks, addition of the diagonal guideline, axis labels, and any theme object created for use with ggplot2 can be supplied.

@@

@@

By default \code{calculate_roc} computes the empirical ROC curve. There are other estimation methods out there, as summarized in the introduction.  Any estimation method can be used, as long as the cutoff, the TPF and the FPF are returned. Then you can simply pass those values in a data frame to the \code{ggplot} function and override the default statistical tranformation to \code{stat_identity}. For example, let us use the binormal method to create a smooth curve. This approach assumes that the test distribution is normal conditional on disease status.

@@

\pkg{plotROC} makes interactive plots by first converting the \code{ggplot} object into a scalable vector graphic (svg) object with the \code{gridSVG::grid.export} function. This function maps each element of the plot to a corresponding element of the svg markup language. Interactivity is then added using \pkg{d3.js} and \proglang{JavaScript} to manipulate those svg elements in response to users' input. The main interactive feature is to display the cutoff labels at the points on the ROC curve closest to the mouse cursor.

There are many ways to solve this with \pkg{d3.js}, but Voronoi polygons are a convenient and efficient way to map the cursor location to approximately the nearest point on the ROC curve. For the set of cutoff points along the ROC curve, the \code{d3.geom.voronoi} function chain computes a set of polygons overlaying the plotting region such that the area of each polygon contains the region of the plot closest to it's corresponding cutoff point. Hover events are bound to the polygons so that when the mouse cursor moves around the plotting region, the closest point on the ROC curve is made visible. Similarly, click events are bound to the polygons so that the appropriate confidence region is made visible upon clicking. The svg code and all necessary \proglang{JavaScript} code is returned in the character string provided by \code{export_interactive_roc}.

This approach is similar to what is done in the \pkg{gridSVG} \code{grid.animate} function, which uses the svg \code{<animate />} tags. However, \pkg{d3.js} provides a much richer set of features. There are several other \proglang{R} packages that aim to create interactive figures. The authors of \pkg{animint} [@animint] created an extensive \proglang{JavaScript} library that creates plots in a similar way as \pkg{ggplot2}. A set of interactive features can be added to plots using \pkg{d3.js}. \pkg{ggvis} [@ggvis], \pkg{rCharts} [@rcharts], and the more recently released \pkg{htmlwidgets} [@htmlwidgets] all leverage existing charting libraries written in \proglang{JavaScript}. \pkg{qtlcharts} [@qtlcharts] uses a set of custom \proglang{JavaScript} and \pkg{d3.js} functions to visualize data from genetic experiments. Their general approach is to manipulate the data and record plotting options in \proglang{R}, and then pass those objects to the charting libraries or functions that handle the rendering and interactivity. \pkg{plotROC} lets \proglang{R} do the rendering, allowing the figures to be consistent across print and web-based media, and retaining the distinctive \proglang{R} style. This also allows users to manipulate the figures directly in \proglang{R} to suit their needs, using tools that are more accessible and familiar to most \proglang{R} users. Then, the \proglang{JavaScript} adds a layer of interactivity to the rendered figures.
